Explore consensus cell types

### 2025-01-23 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd
adjust plots based on review

}

# stacked bar chart showing the distribution of the top 9 cell types for each project, including Unknown
celltype_distribution_plot <- function(label,
                                       results_df,
                                       colors){

  # filter to get only results for the specified project 
    plot_df <- plot_df |> 
      dplyr::filter(project_label == {{label}}) |> 
      dplyr::mutate(
        # relevel factors for specific project 
        top_celltypes = forcats::fct_infreq(top_celltypes) |> 
          forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf)
      )

    # make a stacked bar chart with top cell types 
    ggplot(plot_df) + 
      aes(
        x = library_id, 
        y = percent_cells_annotation, 
        fill = top_celltypes
      ) +
      geom_col() + 
      scale_y_continuous(expand = c(0,0)) +
      scale_fill_manual(values = colors, name = "cell type") +
      ggtitle(label) +
      theme(axis.text.x = element_blank())

@@

```{r}
# list all results files 
results_files <- list.files(results_dir, pattern = "*_consensus-cell-types.tsv.gz", full.names = TRUE)

@@

# read in results and prep data frame for plotting 
all_results_df <- results_files |> 
  purrr::imap(summarize_celltypes) |> 
  dplyr::bind_rows(.id = "project_id") |> 
  # add in diagnoses 
  dplyr::left_join(diagnoses_df, by = "project_id") |> 
  dplyr::mutate(
    # remove leading 0s
    project_string = stringr::str_remove(project_id, "SCPCP0000"),
    # create a label for plotting
    project_label = glue::glue("{project_string}:{diagnosis}")
  )

@@

```{r, fig.height=5}
unknown_only <- all_results_df |> 
  dplyr::filter(consensus_annotation == "Unknown")

ggplot(unknown_only, aes(x = project_label, y = percent_cells_annotation)) +
  ggforce::geom_sina(size = 0.1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5),
        plot.margin = margin(10,10,10,10)) +
  labs(
    x = "", 
    y = "Percent of cells annotated as Unknown"
  )

@@

It looks like we do have some samples that aren't jus all "Unknown"!
It definitely varies by project, but for most projects we at least see some proportion of samples with assigned cell types.

@@

```{r}
high_tumor_df <- unknown_only |> 
  dplyr::mutate(no_cells_identified = dplyr::if_else(percent_cells_annotation == 100, TRUE, FALSE)) |> 
  dplyr::group_by(project_label) |> 
  dplyr::summarize(all_unknown = sum(no_cells_identified),
                   classified_cells = sum(!no_cells_identified),
                   percentage_unknown = round(all_unknown/(all_unknown + classified_cells)*100, 2),
                   # add number of libraries for plotting 
                   total_libraries = length(library_id)) |>
  # set order for plots 
  dplyr::mutate(project_label = forcats::fct_reorder(project_label, total_libraries, .desc = TRUE))
```

@@

```{r, fig.height=5}
# pivot for plotting so above and below count is in the same column 
high_tumor_df <- high_tumor_df |> 
  tidyr::pivot_longer(
    cols = c("all_unknown", "classified_cells"),
    names_to = "category",
    values_to = "number_of_samples"
  )

ggplot(high_tumor_df, aes(x = project_label, y = number_of_samples, fill = category)) +
  geom_bar(position = "stack", stat= "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = rel(0.9))) +
  labs(
    x = "", 
    y = "Number of samples",
    fill = ""
  ) +
  scale_fill_manual(values = c( "#FFCB05", "#00274C"))
```

@@

```{r}
num_celltypes_df <- all_results_df |> 
  # remove unknown as a cell type 
  dplyr::filter(consensus_annotation != "Unknown") |> 
  dplyr::select(project_label, library_id, num_celltypes) |> 
  unique()

ggplot(num_celltypes_df, aes(x = num_celltypes)) +
  geom_histogram(binwidth = 1, center = 0) +
  facet_wrap(project_label)
```

@@

Now we look at the distribution of the cell types in each sample. 
For these plots, we will pull out the top 9 cell types for each project. 
All other cells will be labeld with "All remaining cell types".

The top cell types are determined by counting how many libraries each cell type is found in within a project and taking the most frequent types. 
Note that because the top cell types are different for each project, the colors will be different for assigned cell types. 
The only colors that will stay the same are those for "Unknown" and "All remaining cell types".

@@

```{r}
project_labels <- unique(all_results_df$project_label)

project_labels |> 
  purrr::map(\(label){
    celltype_distribution_plot(label, plot_df, colors)
    })
```

>>>>

@@

```{r}
# list all results files 
results_files <- list.files(results_dir, pattern = "_consensus-cell-types\\.tsv.\\gz$", full.names = TRUE)

@@

# read in results and prep data frame for plotting 
all_results_df <- results_files |> 
  purrr::imap(summarize_celltypes) |> 
  dplyr::bind_rows(.id = "project_id") |> 
  # add in diagnoses 
  dplyr::left_join(diagnoses_df, by = "project_id") |> 
  dplyr::mutate(
    # create a label for plotting
    project_label = glue::glue("{project_id}:{diagnosis}")
  )

@@

```{r, fig.height=7}
unknown_only <- all_results_df |> 
  dplyr::filter(consensus_annotation == "Unknown")

ggplot(unknown_only, aes(x = project_label, y = percent_cells_annotation)) +
  ggforce::geom_sina(size = 0.1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        plot.margin = margin(10,10,10,10)) +
  labs(
    x = "", 
    y = "Percent of cells annotated as Unknown"
  )

@@

It looks like we do have some samples that aren't just all "Unknown"!
It definitely varies by project, but for most projects we at least see some proportion of samples with assigned cell types.

@@

```{r}
high_tumor_df <- unknown_only |> 
  dplyr::mutate(no_cells_identified = percent_cells_annotation == 100) |> 
  dplyr::group_by(project_label) |> 
  dplyr::summarize(all_unknown = sum(no_cells_identified),
                   classified_cells = sum(!no_cells_identified),
                   percentage_unknown = round(all_unknown/(all_unknown + classified_cells)*100, 2),
                   # add number of libraries for plotting 
                   total_libraries = length(library_id)) |>
  # set order for plots 
  dplyr::mutate(project_label = forcats::fct_reorder(project_label, total_libraries, .desc = TRUE))
```

@@

@@

```{r, fig.height=10}
num_celltypes_df <- all_results_df |> 
  # add a new line for facet labels 
  dplyr::mutate(facet_label = glue::glue("{project_id}\n{diagnosis}")) |>
  # remove unknown as a cell type 
  dplyr::filter(consensus_annotation != "Unknown") |> 
  dplyr::select(facet_label, library_id, num_celltypes) |> 
  unique()

ggplot(num_celltypes_df, aes(x = num_celltypes)) +
  geom_histogram(binwidth = 1, center = 0) +
  facet_wrap(vars(facet_label), 
             ncol = 3) +
  labs(
    x = "Number of cell types"
  ) +
  theme_bw()
```

@@

Now we look at the distribution of the cell types in each sample. 
For these plots, we will pull out the top 9 cell types for each project. 
All other cells will be labeled with "All remaining cell types".

The top cell types are determined by counting how many libraries each cell type is found in within a project and taking the most frequent types.

@@

```{r, fig.height=60, fig.width=10}
project_labels <- unique(all_results_df$project_label)

# stacked bar chart showing the distribution of the top 9 cell types for each project, including Unknown
project_labels |> 
  purrr::map(\(label){

    project_df <- plot_df |> 
      dplyr::filter(project_label == {{label}}) |> 
      dplyr::mutate(
        # relevel factors for specific project 
        top_celltypes = forcats::fct_infreq(top_celltypes) |> 
          forcats::fct_relevel("All remaining cell types", "Unknown", after = Inf)
      )

    # make a stacked bar chart with top cell types 
    ggplot(project_df) + 
      aes(
        x = library_id, 
        y = percent_cells_annotation, 
        fill = top_celltypes
      ) +
      geom_col() + 
      scale_y_continuous(expand = c(0,0)) +
      scale_fill_manual(values = colors, name = "cell type") +
      ggtitle(label) +
      theme(axis.text.x = element_blank())

    }) |>
  patchwork::wrap_plots(ncol = 1)
```
